---
title: "White Radakovic Figure 3CD - nutrient starvation"
author: "Laura White"
date: "2024-11-20"
output: html_document
---
```{r global options, include = FALSE}
knitr::opts_chunk$set(message=FALSE)

pkgs <- c("tidyverse",
          "here",
          "cowplot",
          "purrr",
          "ggrepel",
          "FNN")

# load packages
invisible(lapply(pkgs, 
                 library, 
                 character.only = TRUE,
                 warn.conflicts = FALSE,
                 quietly = TRUE))

theme_set(theme_cowplot())

here::i_am("rmd/Figure3CD.Rmd")
```
Define colors for tRNA plotting.
```{r}
# throw in our new color scheme
aa_mapping <- c(
  "Lysine" = "#56B4E9", "Lys" = "#56B4E9", "K" = "#56B4E9",
  "Arginine" = "#0072B2", "Arg" = "#0072B2", "R" = "#0072B2",
  "Aspartic acid" = "#ef3b2c", "Asp" = "#ef3b2c", "D" = "#ef3b2c",
  "Glutamic acid" = "#a50f15", "Glu" = "#a50f15", "E" = "#a50f15",
  "Phenylalanine" = "#d9d9d9", "Phe" = "#d9d9d9", "F" = "#d9d9d9",
  "Tyrosine" = "#737373", "Tyr" = "#737373", "Y" = "#737373",
  "Tryptophan" = "#252525", "Trp" = "#252525", "W" = "#252525",
  "Glycine" = "#c7e9c0", "Gly" = "#c7e9c0", "G" = "#c7e9c0",
  "Alanine" = "#a1d99b", "Ala" = "#a1d99b", "A" = "#a1d99b",
  "Valine" = "#74c476", "Val" = "#74c476", "V" = "#74c476",
  "Isoleucine" = "#238b45", "Ile" = "#238b45", "I" = "#238b45",
  "Leucine" = "#00441b", "Leu" = "#00441b", "L" = "#00441b",
  "Serine" = "#bcbddc", "Ser" = "#bcbddc", "S" = "#bcbddc",
  "Threonine" = "#9e9ac8", "Thr" = "#9e9ac8", "T" = "#9e9ac8",
  "Asparagine" = "#807dba", "Asn" = "#807dba", "N" = "#807dba",
  "Glutamine" = "#4a1486", "Gln" = "#4a1486", "Q" = "#4a1486",
  "Cysteine" = "#fff7bc", "Cys" = "#fff7bc", "C" = "#fff7bc",
  "Histidine" = "#fee391", "His" = "#fee391", "H" = "#fee391",
  "Methionine" = "#fec44f", "Met" = "#fec44f", "iMet" = "#fec44f", "M" = "#fec44f",
  "Proline" = "#fe9929", "Pro" = "#fe9929", "P" = "#fe9929",
  "Selenocysteine" = "#ec7014", "Sec" = "#ec7014", "U" = "#ec7014"
)
```

Pull in CPM-normalized summary tsvs from per-read tsvs produced by running the CCA model on our charged tRNA sequencing data. Can use these to generate abundnace, charging, log2 fold changes in the both.
```{r}
tsv_dir <- here("data/classification_test/leudepletion")

# Get all .tsv file paths in the directory
tsv_files <- list.files(tsv_dir, pattern = "\\.tsv$", full.names = TRUE)

# Import all TSV files into a list & combine into one df with a column including the file name 
tsv_list <- map2(tsv_files, tsv_files, ~ read_tsv(.x) %>% mutate(filename = basename(.y)))

combined_tsv <- bind_rows(tsv_list) %>%
  mutate(filename = str_remove(filename, "\\_charging_likelihood_cpm.tsv$")) %>%  # remove end of file name
  separate(filename, into = c("genotype", "media", "time", "rep")) %>%
  separate(tRNA, into = c("tRNA", "AA", "anticodon", "family", "species")) %>%
  mutate(rep = str_remove(rep, "r")) %>%
  mutate(rep = str_remove(rep, "ep")) %>%
  select(-genotype) # for this one experiment can drop 


combined_summary <- combined_tsv %>%
  group_by(tRNA, AA, anticodon, time, media, rep) %>%
  summarise(
    charged = sum(charged),
    uncharged = sum(uncharged),
    charged_CPM = sum(charged_CPM),
    uncharged_CPM = sum(uncharged_CPM),
    .groups = 'drop'  # Optionally ungroup after summarizing
  )
```

Figure 3 (C) Log2 fold change in tRNA abundance and tRNA charging percent charged reads for all isodecoders after 15 minutes growth in synthetic complete (SC) or leucine dropout media. Points represent the mean of the same 3 biological replicates from panel (B), with error bars spanning the standard deviation.
For our EZ charging northern-matched sequencing data we opted for 15 min timepoints (these are reps 5-7).
```{r}
combined_summary %>%
  select(-charged, -uncharged) %>% # we only care about CPM
  mutate(total = charged_CPM + uncharged_CPM, 
           pct_charged = ifelse(total > 0, 100 * (charged_CPM / total), NA)) -> cleaned_data

# Define the baseline and comparison temperatures
baseline_media <- "SC"
comparison_media <- "minusleu"

# Calculate log2 fold changes
log2fc_data <- cleaned_data %>%
  filter(!str_starts(tRNA, "pre")) %>% # remove pre-tRNAs, low counts on these & dubious
  filter(time == "15m") %>%
  filter(rep %in% c(5, 6, 7)) %>%
  filter(media %in% c(baseline_media, comparison_media)) %>%
  group_by(AA, anticodon, rep) %>%
  summarize(
    log2fc_total = log2(total[media == comparison_media] / total[media == baseline_media]),
    log2fc_pct_charged = log2(pct_charged[media == comparison_media] / pct_charged[media == baseline_media]),
    .groups = 'drop'
  ) %>%
  ungroup() %>%
  mutate(label = paste(AA, anticodon, sep = "-"))  # Create the label column for plotting

repsum_data <- log2fc_data %>%
  group_by(AA, anticodon, label) %>%
  summarize(
    mean_log2fc_total = mean(log2fc_total),
    sd_log2fc_total = sd(log2fc_total),
    mean_log2fc_pct_charged = mean(log2fc_pct_charged),
    sd_log2fc_pct_charged = sd(log2fc_pct_charged),
    .groups = "drop"
  )

ggplot(repsum_data, aes(x = mean_log2fc_total, y = mean_log2fc_pct_charged)) +
    geom_hline(yintercept = 0, color = "black", linetype = "solid", size = 0.5) +  # Darker line at y=0
    geom_vline(xintercept = 0, color = "black", linetype = "solid", size = 0.5) +  # Darker line at x=0
    geom_errorbarh(aes(xmin = mean_log2fc_total - sd_log2fc_total, 
                       xmax = mean_log2fc_total + sd_log2fc_total), 
                   height = 0.2, color = "black", alpha = 0.1) +  # Horizontal error bars
    geom_errorbar(aes(ymin = mean_log2fc_pct_charged - sd_log2fc_pct_charged, 
                      ymax = mean_log2fc_pct_charged + sd_log2fc_pct_charged), 
                  width = 0.2, color = "black", alpha = 0.1) +  # Vertical error bars
    geom_point(aes(color = AA), size = 3) +  # Customize color based on AA, size adjusted for visibility
    labs(
      x = "Log2 fold change (abundance)",
      y = "Log2 fold change (% charging)",
      title = ""
    ) +
    scale_color_manual(values = aa_mapping) +  # Apply your custom color palette
    geom_text_repel(aes(label = label), size = 4) +  # Adjust text size as needed
    theme_cowplot() +
    theme(
          legend.position = "none",
          aspect.ratio = 1,
          plot.background = element_rect(fill = "white", color = NA),
          panel.background = element_rect(fill = "white", color = NA)
          ) +
    coord_fixed(ratio = 1) +
    guides(
      color = guide_legend(nrow = 3, byrow = TRUE)  # Arrange legend into 3 rows, wrapping items across rows
    ) -> plot
ggsave(here("figures", "Leustarvation3C.png"), plot = plot, width = 4, height = 4, units = "in")
  
```

(D) Volcano plot of the mean fold change in aminoacylation for the same 3 replicates in panels (B) and (C), with Z-test p-values on the y-axis and dashed line indicating the 𝛼 threshold.
Z test to find significant changes
```{r}
# Step 1: Calculate mean log2fc_pct_charged for each isodecoder
mean_data <- log2fc_data %>%
  group_by(label, AA) %>%
  summarise(mean_log2fc_pct_charged = mean(log2fc_pct_charged, na.rm = TRUE))

# Step 2: Calculate overall mean and standard deviation across all isodecoder means
overall_mean <- mean(mean_data$mean_log2fc_pct_charged, na.rm = TRUE)
overall_sd <- sd(mean_data$mean_log2fc_pct_charged, na.rm = TRUE)

# Step 3: Calculate Z-scores and corresponding p-values
test_results <- mean_data %>%
  mutate(
    z_score = (mean_log2fc_pct_charged - overall_mean) / overall_sd,
    p_value = 2 * pnorm(-abs(z_score))  # two-sided p-value
  )

# Step 4: Adjust p-values for multiple testing using BH correction
test_results <- test_results %>%
  mutate(adj_p_value = p.adjust(p_value, method = "BH")) %>%
  mutate(xform_adj_pval = -log10(adj_p_value))

# Join the standard deviation for log2fc_pct_charged from repsum_data to test_results
test_results <- test_results %>%
  left_join(repsum_data %>% select(label, AA, sd_log2fc_pct_charged), by = c("label", "AA"))

# plot corrected pvals against log2fc with horizontal error bars
test_results %>%
  ggplot(aes(x = mean_log2fc_pct_charged, y = xform_adj_pval, color = AA, label = label)) +
  geom_hline(yintercept = 1.3, linetype = "dashed", color = "grey50", size = .75) +
  geom_vline(xintercept = 0) +
  geom_errorbarh(aes(xmin = mean_log2fc_pct_charged - sd_log2fc_pct_charged, 
                     xmax = mean_log2fc_pct_charged + sd_log2fc_pct_charged), 
                 height = 0.2, color = "black", alpha = 0.1) +  # Horizontal error bars
  geom_point(size = 3) +  # Plot mean values
  geom_text_repel(aes(label = label), size = 4, color = "black") + 
  scale_color_manual(values = aa_mapping) + 
  labs(
    x = "Log2 fold change (% charging)",
    y = "-log10(p-val)",
    title = "",
    color = "Amino Acid"
  ) +
  theme_cowplot() +
  theme(
        legend.position = "none",
        aspect.ratio = 1,
        plot.background = element_rect(fill = "white", color = NA),
        panel.background = element_rect(fill = "white", color = NA)
        ) +
  coord_fixed(ratio = 1) -> plot
ggsave(here("figures", "Leustarvation3D.png"), plot = plot, width = 4, height = 4, units = "in")
```
