---
title: "White Radakovic Figure 3"
author: "Laura White"
date: "2024-11-20"
output: html_document
---
```{r global options, include = FALSE}
knitr::opts_chunk$set(message=FALSE)

pkgs <- c("tidyverse",
          "here",
          "cowplot",
          "purrr",
          "ggrepel",
          "FNN")

# load packages
invisible(lapply(pkgs, 
                 library, 
                 character.only = TRUE,
                 warn.conflicts = FALSE,
                 quietly = TRUE))

theme_set(theme_cowplot())

here::i_am("rmd/Figure3A.Rmd")
```

Define colors for tRNA plotting.
```{r}
# throw in our new color scheme
aa_mapping <- c(
  "Lysine" = "#56B4E9", "Lys" = "#56B4E9", "K" = "#56B4E9",
  "Arginine" = "#0072B2", "Arg" = "#0072B2", "R" = "#0072B2",
  "Aspartic acid" = "#ef3b2c", "Asp" = "#ef3b2c", "D" = "#ef3b2c",
  "Glutamic acid" = "#a50f15", "Glu" = "#a50f15", "E" = "#a50f15",
  "Phenylalanine" = "#d9d9d9", "Phe" = "#d9d9d9", "F" = "#d9d9d9",
  "Tyrosine" = "#737373", "Tyr" = "#737373", "Y" = "#737373",
  "Tryptophan" = "#252525", "Trp" = "#252525", "W" = "#252525",
  "Glycine" = "#c7e9c0", "Gly" = "#c7e9c0", "G" = "#c7e9c0",
  "Alanine" = "#a1d99b", "Ala" = "#a1d99b", "A" = "#a1d99b",
  "Valine" = "#74c476", "Val" = "#74c476", "V" = "#74c476",
  "Isoleucine" = "#238b45", "Ile" = "#238b45", "I" = "#238b45",
  "Leucine" = "#00441b", "Leu" = "#00441b", "L" = "#00441b",
  "Serine" = "#bcbddc", "Ser" = "#bcbddc", "S" = "#bcbddc",
  "Threonine" = "#9e9ac8", "Thr" = "#9e9ac8", "T" = "#9e9ac8",
  "Asparagine" = "#807dba", "Asn" = "#807dba", "N" = "#807dba",
  "Glutamine" = "#4a1486", "Gln" = "#4a1486", "Q" = "#4a1486",
  "Cysteine" = "#fff7bc", "Cys" = "#fff7bc", "C" = "#fff7bc",
  "Histidine" = "#fee391", "His" = "#fee391", "H" = "#fee391",
  "Methionine" = "#fec44f", "Met" = "#fec44f", "iMet" = "#fec44f", "M" = "#fec44f",
  "Proline" = "#fe9929", "Pro" = "#fe9929", "P" = "#fe9929",
  "Selenocysteine" = "#ec7014", "Sec" = "#ec7014", "U" = "#ec7014"
)
```
Figure 3. Sequencing and analysis of budding yeast tRNAs via chemical-charging northern and aa-tRNA-seq.
A: Correlation of tRNA aminoacylation (% charging) in budding yeast, as measured by acid northern in Fig. 1B (y-axis), and aa-tRNA-seq (mean and standard deviation of 3 replicates)

Here is the densitometry for the acid charging northern with/without chemical ligation.
```{r}
read_csv(here("data/charging_northerns/CN4_densitometry.csv")) %>% 
  rename_at(vars(5), ~"percent_charging") -> cn4

cn4 <- cn4 %>%
  mutate(input = case_when(
    str_detect(Lane, "Acylated") ~ "untreated",
    str_detect(Lane, "Deacylated") ~ "deacylated",
    str_detect(Lane, "50/50") ~ "5050",
    TRUE ~ NA_character_  # This will assign NA for rows that don't match any of the conditions\
  ))

cn4 <- cn4 %>%
  mutate(prep = case_when(
    str_detect(Lane, "input") ~ "input",
    str_detect(Lane, "-") ~ "unligated",
    str_detect(Lane, "\\+") ~ "chemligated",
    TRUE ~ NA_character_  # This will assign NA for rows that don't match any of the conditions\
  ))

cn4 %>%
  filter(prep == "chemligated" & input == "untreated") %>%
  select(-Lane, -prep, -input) %>%
  filter(!AA %in% c("Pro", "Glu", "Ser", "Asp")) %>%
  pivot_wider(names_from = Band, values_from = percent_charging, values_fill = list(percent_charging = 0)) %>%
  rename_with(~ "intermediate", matches("\\*")) %>%
  rename(acylated =  Acylated) %>%
  mutate(ligated = `Chemi Ligated` + `Chemi Ligated 1` + `Chemi Ligated 2`) %>%
  mutate(deacylated = `Deacylated` + `Deacylated 1` + `Deacylated 2`) %>%
  select(AA, anticodon, deacylated, acylated, intermediate, ligated) %>%
  mutate(pct_fully_ligated = (ligated / (acylated + intermediate + ligated)) * 100) %>%
  mutate(pct_shifted = ((ligated + intermediate) / (acylated + intermediate + ligated)) * 100) %>%
  mutate(AA_anticodon = paste(AA, anticodon, sep = "-")) -> cleaned_subset
```

Here we read in in the Remora-classified percent charging data for our wild type yeast samples (without chemical deacylation), and tidy this data.
```{r}
remora <- read_tsv(here("data/2024_chemicaldeacylation/percent_charged_cca_model.tsv"))

remora %>%
  select(tRNA, unt1_percent_charg, unt2_percent_charg, unt3_percent_charg) %>%
  pivot_longer(cols = starts_with("unt"),
               names_to = "replicate",
               values_to = "percent_charged") %>%
  mutate(replicate = case_when(
    replicate == "unt1_percent_charg" ~ "rep1",
    replicate == "unt2_percent_charg" ~ "rep2",
    replicate == "unt3_percent_charg" ~ "rep3"
  )) %>%
  rename(AA_anticodon = tRNA) -> remora_long
```
```{r}
# Step 1: Calculate the mean and SE for percent_charged in wtsc_pct_seqed_summarized
mean_sequencing_data <- remora_long %>%
  mutate(AA_anticodon = str_replace(AA_anticodon, "_", "-")) %>%
  group_by(AA_anticodon) %>%
  summarise(
    mean_percent_charged = mean(percent_charged, na.rm = TRUE),
    se_percent_charged = sd(percent_charged, na.rm = TRUE) / sqrt(n())  # SE = SD / sqrt(n)
  )

# Step 2: Join with northern data (test)
cleaned_subset %>%
  select(AA_anticodon, deacylated) %>%
  mutate(ncharged = 100 - deacylated) -> test

# Perform the join and separate AA and anticodon
merged_data <- mean_sequencing_data %>%
  ungroup() %>%
  separate(AA_anticodon, into = c("AA", "anticodon"), sep = "-", remove = FALSE) %>%
  left_join(test, by = "AA_anticodon") %>%
  filter(complete.cases(.))


# Step 4: Perform the correlation test
cor_test <- cor.test(merged_data$ncharged, merged_data$mean_percent_charged, use = "complete.obs")

# Extract the correlation coefficient and p-value
correlation <- cor_test$estimate
p_value <- cor_test$p.value

# Step 5: Create the scatter plot with error bars for SE
ggplot(merged_data, aes(x = mean_percent_charged, y = ncharged)) +
  geom_abline(intercept = 0, slope = 1, color = "grey90", linetype = "solid") +
  geom_smooth(method = "lm", se = FALSE, color = "gray", linetype = "dashed") +
  geom_errorbar(aes(xmin = mean_percent_charged - se_percent_charged, xmax = mean_percent_charged + se_percent_charged), color = "gray50", width = 0.2) +
  geom_point(aes(color = AA), size = 3) +
  geom_text_repel(aes(label = AA_anticodon)) +
  scale_color_manual(values = aa_mapping) +
  labs(
    y = "% charging (northern)",
    x = "% charging (sequencing)",
    title = "",
    color = "", shape = ""
  ) +
  xlim(0, 80) +  # Adjust axis limits as needed
  ylim(0, 80) +  # Adjust axis limits as needed
  theme_cowplot() +
  coord_fixed(ratio = 1) +
  theme(legend.position = "none",
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),) +
  # Add correlation coefficient and p-value as annotations
  annotate("text", y = 15, x = 40, label = paste("r =", round(correlation, 2)), size = 5, hjust = 0) +
  annotate("text", y = 10, x = 40, label = paste("p-val = ", format.pval(p_value, digits = 3, scientific = T)), size = 5, hjust = 0) -> plot


# Now try saving the plot
ggsave(here("figures", "Northerndensitometry3A.png"), plot = plot, width = 4, height = 4, units = "in")
```
