---
title: "White Radakovic Figure 3"
author: "Laura White"
date: "2024-11-20"
output: html_document
---
```{r global options, include = FALSE}
knitr::opts_chunk$set(message=FALSE)

pkgs <- c("tidyverse",
          "here",
          "cowplot",
          "purrr",
          "ggrepel",
          "FNN")

# load packages
invisible(lapply(pkgs, 
                 library, 
                 character.only = TRUE,
                 warn.conflicts = FALSE,
                 quietly = TRUE))

theme_set(theme_cowplot())

here::i_am("rmd/Figure3.Rmd")
```

Define colors for tRNA plotting.
```{r}
# throw in our new color scheme
aa_mapping <- c(
  "Lysine" = "#56B4E9", "Lys" = "#56B4E9", "K" = "#56B4E9",
  "Arginine" = "#0072B2", "Arg" = "#0072B2", "R" = "#0072B2",
  "Aspartic acid" = "#ef3b2c", "Asp" = "#ef3b2c", "D" = "#ef3b2c",
  "Glutamic acid" = "#a50f15", "Glu" = "#a50f15", "E" = "#a50f15",
  "Phenylalanine" = "#d9d9d9", "Phe" = "#d9d9d9", "F" = "#d9d9d9",
  "Tyrosine" = "#737373", "Tyr" = "#737373", "Y" = "#737373",
  "Tryptophan" = "#252525", "Trp" = "#252525", "W" = "#252525",
  "Glycine" = "#c7e9c0", "Gly" = "#c7e9c0", "G" = "#c7e9c0",
  "Alanine" = "#a1d99b", "Ala" = "#a1d99b", "A" = "#a1d99b",
  "Valine" = "#74c476", "Val" = "#74c476", "V" = "#74c476",
  "Isoleucine" = "#238b45", "Ile" = "#238b45", "I" = "#238b45",
  "Leucine" = "#00441b", "Leu" = "#00441b", "L" = "#00441b",
  "Serine" = "#bcbddc", "Ser" = "#bcbddc", "S" = "#bcbddc",
  "Threonine" = "#9e9ac8", "Thr" = "#9e9ac8", "T" = "#9e9ac8",
  "Asparagine" = "#807dba", "Asn" = "#807dba", "N" = "#807dba",
  "Glutamine" = "#4a1486", "Gln" = "#4a1486", "Q" = "#4a1486",
  "Cysteine" = "#fff7bc", "Cys" = "#fff7bc", "C" = "#fff7bc",
  "Histidine" = "#fee391", "His" = "#fee391", "H" = "#fee391",
  "Methionine" = "#fec44f", "Met" = "#fec44f", "iMet" = "#fec44f", "M" = "#fec44f",
  "Proline" = "#fe9929", "Pro" = "#fe9929", "P" = "#fe9929",
  "Selenocysteine" = "#ec7014", "Sec" = "#ec7014", "U" = "#ec7014"
)
```
Figure 3. Sequencing and analysis of budding yeast tRNAs via chemical-charging northern and aa-tRNA-seq.

Here is the densitometry for the acid charging northern with/without chemical ligation.
```{r}
read_csv(here("data/charging_northerns/CN4_densitometry.csv")) %>% 
  rename_at(vars(5), ~"percent_charging") -> cn4

cn4 <- cn4 %>%
  mutate(input = case_when(
    str_detect(Lane, "Acylated") ~ "untreated",
    str_detect(Lane, "Deacylated") ~ "deacylated",
    str_detect(Lane, "50/50") ~ "5050",
    TRUE ~ NA_character_  # This will assign NA for rows that don't match any of the conditions\
  ))

cn4 <- cn4 %>%
  mutate(prep = case_when(
    str_detect(Lane, "input") ~ "input",
    str_detect(Lane, "-") ~ "unligated",
    str_detect(Lane, "\\+") ~ "chemligated",
    TRUE ~ NA_character_  # This will assign NA for rows that don't match any of the conditions\
  ))

cn4 %>%
  filter(prep == "chemligated" & input == "untreated") %>%
  select(-Lane, -prep, -input) %>%
  filter(!AA %in% c("Pro", "Glu", "Ser", "Asp")) %>%
  pivot_wider(names_from = Band, values_from = percent_charging, values_fill = list(percent_charging = 0)) %>%
  rename_with(~ "intermediate", matches("\\*")) %>%
  rename(acylated =  Acylated) %>%
  mutate(ligated = `Chemi Ligated` + `Chemi Ligated 1` + `Chemi Ligated 2`) %>%
  mutate(deacylated = `Deacylated` + `Deacylated 1` + `Deacylated 2`) %>%
  select(AA, anticodon, deacylated, acylated, intermediate, ligated) %>%
  mutate(pct_fully_ligated = (ligated / (acylated + intermediate + ligated)) * 100) %>%
  mutate(pct_shifted = ((ligated + intermediate) / (acylated + intermediate + ligated)) * 100) %>%
  mutate(AA_anticodon = paste(AA, anticodon, sep = "-")) -> cleaned_subset
```

Here we read in in the Remora-classified percent charging data for our wild type yeast samples (without chemical deacylation), and tidy this data.
```{r}
remora <- read_tsv(here("data/2024_chemicaldeacylation/percent_charged_cca_model.tsv"))

remora %>%
  select(tRNA, unt1_percent_charg, unt2_percent_charg, unt3_percent_charg) %>%
  pivot_longer(cols = starts_with("unt"),
               names_to = "replicate",
               values_to = "percent_charged") %>%
  mutate(replicate = case_when(
    replicate == "unt1_percent_charg" ~ "rep1",
    replicate == "unt2_percent_charg" ~ "rep2",
    replicate == "unt3_percent_charg" ~ "rep3"
  )) %>%
  rename(AA_anticodon = tRNA) -> remora_long

remora_old %>%
  select(tRNA, percent_charg_untr1, percent_charg_untr2, percent_charg_untr3) %>%
  pivot_longer(cols = starts_with("per"),
               names_to = "replicate",
               values_to = "percent_charged") %>%
  mutate(replicate = case_when(
    replicate == "percent_charg_untr1" ~ "rep1",
    replicate == "percent_charg_untr2" ~ "rep2",
    replicate == "percent_charg_untr3" ~ "rep3"
  )) %>%
  rename(AA_anticodon = tRNA) -> remora_old_long
```

