---
title: "White Radakovic Figure 4, S4, S5 - synthetic tRNA analysis"
author: "Laura White"
date: "2024-11-20"
output: html_document
---
```{r global options, include = FALSE}
knitr::opts_chunk$set(message=FALSE)

pkgs <- c("tidyverse",
          "here",
          "cowplot",
          "purrr",
          "forcats",
          "ggrepel",
          "magrittr",
          "scales",
          "viridis")

# load packages
invisible(lapply(pkgs, 
                 library, 
                 character.only = TRUE,
                 warn.conflicts = FALSE,
                 quietly = TRUE))

theme_set(theme_cowplot())

aa_mapping <- c(
  "Lysine" = "#56B4E9", "Lys" = "#56B4E9", "K" = "#56B4E9",
  "Arginine" = "#0072B2", "Arg" = "#0072B2", "R" = "#0072B2",
  "Aspartic acid" = "#ef3b2c", "Asp" = "#ef3b2c", "D" = "#ef3b2c",
  "Glutamic acid" = "#a50f15", "Glu" = "#a50f15", "E" = "#a50f15",
  "Phenylalanine" = "#d9d9d9", "Phe" = "#d9d9d9", "F" = "#d9d9d9",
  "Tyrosine" = "#737373", "Tyr" = "#737373", "Y" = "#737373",
  "Tryptophan" = "#252525", "Trp" = "#252525", "W" = "#252525",
  "Glycine" = "#c7e9c0", "Gly" = "#c7e9c0", "G" = "#c7e9c0",
  "Alanine" = "#a1d99b", "Ala" = "#a1d99b", "A" = "#a1d99b",
  "Valine" = "#74c476", "Val" = "#74c476", "V" = "#74c476",
  "Isoleucine" = "#238b45", "Ile" = "#238b45", "I" = "#238b45",
  "Leucine" = "#00441b", "Leu" = "#00441b", "L" = "#00441b",
  "Serine" = "#bcbddc", "Ser" = "#bcbddc", "S" = "#bcbddc",
  "Threonine" = "#9e9ac8", "Thr" = "#9e9ac8", "T" = "#9e9ac8",
  "Asparagine" = "#807dba", "Asn" = "#807dba", "N" = "#807dba",
  "Glutamine" = "#4a1486", "Gln" = "#4a1486", "Q" = "#4a1486",
  "Cysteine" = "#fff7bc", "Cys" = "#fff7bc", "C" = "#fff7bc",
  "Histidine" = "#fee391", "His" = "#fee391", "H" = "#fee391",
  "Methionine" = "#fec44f", "Met" = "#fec44f", "iMet" = "#fec44f", "M" = "#fec44f",
  "Proline" = "#fe9929", "Pro" = "#fe9929", "P" = "#fe9929",
  "Selenocysteine" = "#ec7014", "Sec" = "#ec7014", "U" = "#ec7014"
)

here::i_am("rmd/Figure4_S4_S5.Rmd")

```
Take summary metrics extracted from Remora for synthetic aa-tRNA analysis.
```{r}
tsv_dir <- here("data/remora_signals")
tsv_files <- list.files(tsv_dir, pattern = "\\.tsv.gz$", full.names = TRUE)


tsv_list <- map(tsv_files, ~ read_tsv(.x, skip = 1) %>%
  mutate(filename = basename(.x)) %>%
  group_by(Metric, Reference_Position, filename) %>%
  summarise(mean = mean(Value, na.rm = TRUE), 
            sd = sd(Value, na.rm = TRUE), .groups = 'drop')
)

combined_summary <- bind_rows(tsv_list)

combined_summary <- combined_summary %>%
  mutate(sample = str_replace(basename(filename), "\\.tsv\\.gz$", ""), 
         Reference_Position = Reference_Position - 24) %>%  # Adjust reference position for adapters
  select(-filename)  # Drop the original filename column
```
Make a plot function to generate plots of two samples against each other.
```{r}
# Define the plotting function
plot_two_samples <- function(data, sample1, sample2, metric_filter) {
  # Filter data for the two specified samples and the chosen metric
  filtered_data <- data %>%
    filter(sample %in% c(sample1, sample2), Metric == metric_filter) %>%
    mutate(sample = factor(sample, levels = c(sample2, sample1)))  # Ensure sample order
  
  # Define color mapping to ensure "uncharged" is always blue and the other sample is red
  color_mapping <- setNames(c("#56B4E9", "#D55E00"), c(sample1, sample2))
  
  # Set y-axis label based on the metric
  y_axis_label <- ifelse(metric_filter == "trimmean", "Mean Current (pA)", "Mean Dwell (ms)")
  
  # Generate the plot
  p <- ggplot(filtered_data, aes(x = Reference_Position, y = mean, color = sample, fill = sample)) +
    geom_line(size = 1) +  # Line plot for the mean
    geom_ribbon(aes(ymin = mean - sd, ymax = mean + sd), alpha = 0.3, linetype = 0) +
    scale_color_manual(values = color_mapping) +  # Fixed color mapping
    scale_fill_manual(values = color_mapping) +
    theme_cowplot() +
    geom_vline(xintercept = 101 - 24, linetype = "dashed", size = 1.2) +  # Adjusted reference position and thicker dashed line
    labs(title = "",
         x = "Position",
         y = y_axis_label) +  # Dynamically set y-axis label
    scale_x_continuous(breaks = seq(0, max(filtered_data$Reference_Position), by = 5),  # Major ticks every 5
                       minor_breaks = seq(0, max(filtered_data$Reference_Position), by = 1)) +  # Minor ticks every 1
    theme(
      legend.position = c(0.05, 0.95),  # Position legend in the top left of the plot area
      legend.justification = c("left", "top"),  # Align legend to top left corner
      legend.background = element_rect(fill = "white", color = NA, size = 0.5),  # Optional: add background to legend
      legend.title = element_blank(),
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    )
  
  return(p)
}
```
Now work through all of them. These represent the panels in supplemental figures 4 & 5 in the preprint.
```{r}
# Define the metrics to be plotted
metrics <- c("trimmean", "dwells")

# Get a list of unique sample names, excluding "uncharged"
all_samples <- unique(combined_summary$sample)
comparison_samples <- setdiff(all_samples, "uncharged")

# Loop through each metric, sample, and generate plots against "uncharged", saving each plot
for (metric in metrics) {
  for (sample in comparison_samples) {
    # Generate the plot for the current sample compared to "uncharged" with the specified metric
    plot <- plot_two_samples(combined_summary, "uncharged", sample, metric_filter = metric)
    
    # Define the file name based on the metric and sample name
    file_name <- paste0("SuppSynthetic.", metric, "_uncharged_vs_", sample, ".png")
    
    # Save the plot to the specified directory
    ggsave(here("figures", file_name), plot = plot, width = 7, height = 3, units = "in")
    
    # Optional: Print a message to confirm each plot is saved
    message("Saved plot for ", metric, " uncharged vs ", sample, " as ", file_name)
  }
}
```
Make heatmap plots for Figure 4A, B
```{r}
amino_acids <- data.frame(
  AA = c("A", "C", "D", "E", "F", "G", "H", "I", "K", "L", "M", "N", "P", "Q", "R", "S", "T", "V", "W", "Y"),
  AA_3letter = c("Ala", "Cys", "Asp", "Glu", "Phe", "Gly", "His", "Ile", "Lys", "Leu", "Met", "Asn", "Pro", "Gln", "Arg", "Ser", "Thr", "Val", "Trp", "Tyr"),
  Charge = c(0, 0, -1, -1, 0, 0, 0, 0, 0.2, 0, 0, 0, 0, 0, 0.2, 0, 0, 0, 0, 0),
  Volume = c(60.4, 73.4, 73.8, 85.9, 121.2, 43.2, 98.8, 107.5, 108.5, 107.5, 105.3, 78, 81, 93.9, 127.3, 60.3, 76.8, 90.8, 143.9, 123.1),
  MolecularWeight = c(89, 121, 133, 147, 165, 75, 155, 131, 146, 131, 149, 132, 115, 146, 174, 105, 119, 117, 204, 181),
  # Adding Hydrophobicity Index columns
  Hydrophobicity_pH2 = c(47, 52, -18, 8, 100, 0, -42, 100, -37, 100, 74, -41, -46, -18, -26, -7, 13, 79, 84, 49),
  Hydrophobicity_pH7 = c(41, 49, -55, -31, 100, 0, 8, 99, -23, 97, 74, -28, -46, -10, -14, -5, 13, 76, 97, 63)
)

# Convert sample to initial caps
combined_summary_cleaned <- combined_summary %>%
  mutate(sample = str_to_title(sample))  # Convert to initial caps

# Join with amino_acids
combined_summary_ordered <- combined_summary_cleaned %>%
  left_join(amino_acids, by = c("sample" = "AA_3letter"))

# Define the exact order for the sample factor levels
custom_order <- rev(c("Trp", "Tyr", "Arg", "Phe", "His", "Met", "Glu", 
                      "Lys", "Gln", "Asp", "Asn", "Ile", "Leu", "Cys", 
                      "Thr", "Val", "Pro", "Ser", "Ala", "Gly", "Uncharged"))

# Apply the custom factor levels to the sample column
combined_summary_ordered <- combined_summary_ordered %>%
  mutate(sample = factor(sample, levels = custom_order))

ggplot(combined_summary_ordered %>% filter(Metric == "dwells"), 
       aes(x = Reference_Position, y = sample, fill = mean)) +
  geom_tile(color = "white") +
  geom_vline(xintercept = 101 - 24, linetype = "dashed", size = 0.5) +
  scale_x_continuous(breaks = seq(0, max(combined_summary_ordered$Reference_Position), by = 5)) +
  scale_fill_viridis_c(
    option = "rocket", direction = -1, 
    limits = c(0, 1200),
    breaks = seq(0, 1200, by = 200),  # More frequent breaks every 200 units
    guide = guide_colorbar(
      title = NULL,                # Remove the legend title
      barwidth = unit(3, "in"),    # Set a fixed width for horizontal layout
      barheight = unit(0.1, "in"), # Ensure bar height remains short for a horizontal layout
      direction = "horizontal",    # Force horizontal orientation
      frame.colour = "black",      # Add a border to the color bar
      frame.linewidth = 0.5,       # Set the border width
      ticks.colour = "black"       # Set tick color for contrast
    )
  ) +
  labs(title = "Mean dwell time (milliseconds)",
       x = "",
       y = "") +
  theme_cowplot() +
  theme(
    legend.position = "top",               # Place the legend at the top
    legend.justification = "center",       # Center the legend
    axis.text.x = element_text(angle = 0, vjust = 0.5, margin = margin(t = 2)),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 10),
    legend.text = element_text(size = 8)   # Adjust legend text size as needed
  ) +
  coord_fixed() # keep tiles square

ggsave(here("figures", "Figure4A.png"), width = 5.5, height = 4, units = "in")
```
same for dwell
```{r}
# Step 1: Specify the exact order for the sample factor levels with "Uncharged" at the bottom
custom_order <- rev(c("Trp", "Tyr", "Arg", "Phe", "His", "Met", "Glu", 
                      "Lys", "Gln", "Asp", "Asn", "Ile", "Leu", "Cys", 
                      "Thr", "Val", "Pro", "Ser", "Ala", "Gly", "Uncharged"))

# Filter data for "trimmean" metric and apply custom order to sample factor
combined_summary_trimmean <- combined_summary %>%
  filter(Metric == "trimmean") %>%
  mutate(sample = factor(str_to_title(sample), levels = custom_order))

# Identify the reference values for the "Uncharged" sample for each position
reference_values <- combined_summary_trimmean %>%
  filter(sample == "Uncharged") %>%
  select(Reference_Position, mean) %>%
  rename(reference_mean = mean)

# Step 2: Join the reference values with the main dataset and calculate charged minus uncharged difference
combined_rel_diff <- combined_summary_trimmean %>%
  left_join(reference_values, by = "Reference_Position") %>%
  mutate(
    diff = mean - reference_mean  # Calculate difference without taking absolute value
  )

# Step 3: Plot the heatmap for the mean difference
ggplot(combined_rel_diff, 
       aes(x = Reference_Position, y = sample, fill = diff)) +
  geom_tile(color = "white") +
  geom_vline(xintercept = 101 - 24, linetype = "dashed", size = 0.5) +
  scale_x_continuous(breaks = seq(0, max(combined_rel_diff$Reference_Position), by = 5)) +
  scale_fill_gradient2(
    name = NULL,
    low = "#0072B2", mid = "white", high = "#D55E00", midpoint = 0,
    limits = c(-0.8, 0.8),                 # Set limits for the difference values
    breaks = seq(-0.8, 0.8, by = 0.2),
    guide = guide_colorbar(
      barwidth = unit(3, "in"),    
      barheight = unit(0.1, "in"), 
      direction = "horizontal",    
      frame.colour = "black",      
      frame.linewidth = 0.5,       
      ticks.colour = "black"       
    )
  ) +
  labs(title = "Relative difference in normalized current (pA)",
       x = "",
       y = "") +
  theme_cowplot() +
  theme(
    title = NULL,
    legend.position = "top",
    legend.justification = "center",
    axis.text.x = element_text(angle = 0, vjust = 0.5, margin = margin(t = 2)),
    axis.text.y = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 10),
    legend.text = element_text(size = 8)
  ) +
  coord_fixed() # Keep tiles square
ggsave(here("figures", "Figure4B.png"), width = 5.5, height = 4, units = "in")
```
Panels with AA features, 4C-4H
Plot correlation between molecular weight of AAs and the mean signals at the largest dwell, current distortions (nts 77, 86)
```{r}
# Join combined_rel_diff with amino_acids and filter for Reference_Position == "77"
rel_diff_joined <- combined_rel_diff %>%
  left_join(amino_acids, by = c("sample" = "AA_3letter")) %>%
  filter(Reference_Position == "77", !is.na(diff))  # Filter for specific reference position and exclude NA values

# Helper function to calculate correlation and p-value
calc_cor_info <- function(data, x_var) {
  model <- lm(diff ~ get(x_var), data = data)
  list(
    estimate = cor(data$diff, data[[x_var]], use = "complete.obs"),
    pvalue = summary(model)$coefficients[2, 4]
  )
}

# Helper function for creating a plot
plot_rel_diff <- function(data, x_var, x_label, cor_info) {
  ggplot(data, aes_string(x = x_var, y = "diff", color = "sample")) +
    geom_smooth(method = "lm", se = FALSE, color = "grey80", linetype = "dashed") +
    geom_point(size = 3) +
    geom_text_repel(aes(label = AA), size = 4, color = "black", max.overlaps = Inf) +
    labs(x = x_label, y = "Current difference (pA)") +
    scale_color_manual(values = aa_mapping) +
    theme_cowplot() +
    theme(legend.position = "none", aspect.ratio = 1, plot.background = element_rect(fill = "white")) +
    scale_y_continuous(limits = c(-0.8, 0.4), breaks = seq(-0.8, 0.4, by = 0.2)) +
    # Place correlation annotation in the top right corner
    annotate("text", x = Inf, y = Inf, 
             label = paste("r =", round(cor_info$estimate, 2), "\np =", format.pval(cor_info$pvalue, digits = 3)),
             hjust = 1.2, vjust = 1.2, size = 4, color = "black")
}


# Step 2: Calculate correlation and create plots
# Molecular Weight
cor_info_mw <- calc_cor_info(rel_diff_joined, "MolecularWeight")
p1_rel_diff <- plot_rel_diff(rel_diff_joined, "MolecularWeight", "Molecular Weight", cor_info_mw)

# Volume
cor_info_vol <- calc_cor_info(rel_diff_joined, "Volume")
p2_rel_diff <- plot_rel_diff(rel_diff_joined, "Volume", "Volume", cor_info_vol)

# Hydrophobicity at pH 7
cor_info_hydro <- calc_cor_info(rel_diff_joined, "Hydrophobicity_pH7")
p3_rel_diff <- plot_rel_diff(rel_diff_joined, "Hydrophobicity_pH7", "Hydrophobicity (pH 7)", cor_info_hydro)

# Save plots with a white background
ggsave(here("figures", "4G.currentbymw.png"), plot = p1_rel_diff, width = 3, height = 3, units = "in", bg = "white")
ggsave(here("figures", "4F.currentbyvolume.png"), plot = p2_rel_diff, width = 3, height = 3, units = "in", bg = "white")
ggsave(here("figures", "4H.currentbyhydrophobicity.png"), plot = p3_rel_diff, width = 3, height = 3, units = "in", bg = "white")

```

```{r}
# Helper function for calculating correlation and p-value
calc_cor <- function(data, x_var) {
  model <- lm(mean ~ get(x_var), data = data)
  list(
    estimate = cor(data$mean, data[[x_var]], use = "complete.obs"),
    pvalue = summary(model)$coefficients[2, 4]
  )
}

# Helper function for plotting
plot_dwells <- function(data, x_var, x_label, cor_info) {
  ggplot(data, aes_string(x = x_var, y = "mean", color = "sample")) +
    geom_smooth(method = "lm", se = FALSE, color = "grey80", linetype = "dashed") +
    geom_point(size = 3) +
    geom_text_repel(aes(label = AA), size = 4, color = "black", max.overlaps = Inf) +  # Repel labels
    labs(x = x_label, y = "Dwell time (ms)") +
    scale_color_manual(values = aa_mapping) +
    theme_cowplot() +
    theme(legend.position = "none", aspect.ratio = 1, panel.background = element_rect(fill = "white")) +
    scale_y_continuous(limits = c(0, 1200)) +
    # Add correlation annotation with relative placement
    annotate("text", x = Inf, y = -Inf, 
             label = paste("r =", round(cor_info$estimate, 2), "\np =", format.pval(cor_info$pvalue, digits = 3)),
             hjust = 1.2, vjust = -0.2, size = 4, color = "black")
}

# Step 1: Filter data for dwells at Reference_Position == 86
filtered_dwells <- combined_summary_ordered %>%
  filter(Metric == "dwells", Reference_Position == 86)

# Step 2: Calculate correlations and p-values
cor_mw_info <- calc_cor(filtered_dwells, "MolecularWeight")
cor_vol_info <- calc_cor(filtered_dwells, "Volume")
cor_hydro_info <- calc_cor(filtered_dwells, "Hydrophobicity_pH7")

# Step 3: Create plots
p1_dwells <- plot_dwells(filtered_dwells, "MolecularWeight", "Molecular Weight", cor_mw_info)
p2_dwells <- plot_dwells(filtered_dwells, "Volume", "Volume", cor_vol_info)
p3_dwells <- plot_dwells(filtered_dwells, "Hydrophobicity_pH7", "Hydrophobicity (pH 7)", cor_hydro_info)

# Step 4: Save plots with a white background
ggsave(here("figures", "4D.dwellbymw.png"), plot = p1_dwells, width = 3, height = 3, units = "in", bg = "white")
ggsave(here("figures", "4C.dwellbyvolume.png"), plot = p2_dwells, width = 3, height = 3, units = "in", bg = "white")
ggsave(here("figures", "4E.dwellbyhydrophobicity.png"), plot = p3_dwells, width = 3, height = 3, units = "in", bg = "white")
```
